Unprepared colonoscopy identifies microbiota specific to the proximal and distal human colon

### 2017-07-27 submission/manuscript.Rmd
updated inline values and citations for v2

```{r oncogenic_otus, echo=FALSE}
subsampled_meta <- merge(meta_file, subsample, by.x='group', by.y='row.names')
fuso179 <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient", "location", "Otu00179")]

@@

p_152[,4] <- (p_152[,3]/subsampled_to) *100
names(p_152)[4] <- "Otu152_abund"
```

@@

Next, we built a model to differentiate the proximal and distal lumenal samples. The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, `r round(cv10f_roc_muc$auc,3)`) compared to the proximal versus distal lumen (`r round(cv10f_roc_lum$auc, 3)`). These models were able to explain ((X%)) of the variance, respectively. OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B). Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C). This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also helped to differentiate between the two sides- such as _Anaerococcus_ and _Finegoldia_.

#### Bacterial OTUs associated with cancer are found in healthy individuals 
Given that specific bacterial species have been associated with colorectal cancer and IBD, we probed our sample set for these OTUs. Among our 100 samples, the most frequent sequence associated with the _Fusobacterium_ genus was OTU179, which aligns via BLASTn to _Fusobacterium nucleatum subsp animalis_ (XX% over full length). This is the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. (@Castellarin2011). The _Fusobacterium_ positive samples were located in x% of the the proximal and X% of the distal mucosa and represented as much as 1% of any sample (Figure 6A). OTU152 was similar to the members of the _Porphyromonas_ genus and the most frequent sequence in that OTU aligned to _Porphyromonas asacharolytica_ (X% over full length), another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Figure 6B). Among the samples that were positive for _P. asacharolytica_, the relative abundances for this OTU ranged from 0.01% - 16%. Thus, disease-associated OTUs could be found in our sample set of 20 healthy individuals.

@@

To get around the noisiness from a diverse set of samples, we built a Random Forest model to identify microbes specific to each side. For each comparison we identified top X OTUs that were strongly predictive of one site or another. Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis @Donaldson2015. For instance, the proximal mucosa contains the highest oxygen concentratons of the colon and  harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and aerobic _Psuedomonas_. The distal mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_. The model was less effective at classifying the proximal and distal lumenal contents, probably because the samples are arguably composed of the same bacteria but differ in water content.

We detected _F. nucleatum_ and _P. asacharolytica_ in 8 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans @Baxter2016 and have oncogenic properties in cell culture and in mice @Sears2014. Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the distal mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous distal mucosa (@Albenberg2014). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on proximal-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (@Mima2016).  Of the 8 (40%) individuals positive for _F. nucleatum_ in this study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa. The presence of _F. nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers (cite). Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis (@Dejea2014). Finally, _Fusobacterium_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model (@Whitmore2014, @Flynn2016). Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of IBD or CRC subtypes in the proximal or distal colon.

The _Fusobacterium_ species _nucleatum_ and _varium_ have been commonly isolated from mucosal biopsies of patients with IBD (cite). Laboratory experiments with these isolates have shown that disease-isolated _F. nucleatum_ are more invasive and stimulate more TNF-a production than strains from healthy individuals (@Strauss2011), suggesting the bacteria may increase inflammation in the gut as well (cite - Dharmani 2011). _F. varium_ isolated from UC patients caused colonic ulcers in an experimental mouse model (Ohkusa 2003). _F. varium_ was only detected in 3 of our study participants and two of those samples were isolated form the proximal mucosa (cite figure). Although there is low occurrance in our study, _F. varium_ is most commonly isolated from UC patient biopsies from the ileum or cecum (cite), suggesting this species may exhibit preference for the different environmental conditions of these gastrointestinal sites. Further work will assess how gut environment may select for species which then cause localized disease.

Specific comparisons of our findings to previously published gut biogeography studies are additionally confounded by the use of bowel preparation methods in most other studies. A rare report of a matched-colonoscopy study sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing (@Shobar2016). This group found that mucosal and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site (@Shobar2016). Bowel cleansing not only made the samples harder to distinguish, it resulted in decreases in diversity across sites. Further, the differences were not great enough to overcome interpersonal differences between subjects. Bowel preparation clearly induces bias into the microbes recovered from sampling the lumen or mucosa of a prepared bowel. Thus our findings in this study are strengthened by the lack of bowel preparation.

@@

#### Sample processing, sequencing and analysis
DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples. The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described (@Kozich2013). Sequences were curated using the mothur software as described previously (@Schloss2009). The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) (@Wang2007) and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff. Sequencing and analysis of a mock community revealed the error rate to be X%. Samples were rarefied to 4231 sequences per sample in order to reduce uneven sampling bias.

Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur version <3.2?> (@Schloss2009). ThetaYC distances were calculated to determine the dissimilarity between two samples. Random Forest classification models were built using the randomForest R package and resultant models were used to identify the OTUs that were most important for classifying each location (@Liaw2002). To get species-level information about sequences of interest, sequences were aligned using blastn and the species name was only used if the identity score was >= 99%.

@@

####Figure 6
Location and abundance of cancer-associated OTUs. Relative abundance was calculated and plotted by sample site for each OTU of interest: (A) _Fusobacterium nucleatum_ and (B) _Porphyromonas asacharolytica_

>>>>

```{r oncogenic_otus, echo=FALSE}
subsampled_meta <- merge(meta_file, subsampled_file, by.x='group', by.y='row.names')
fuso179 <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient", "location", "Otu00179")]
subsampled_to = 4231

@@

p_152[,4] <- (p_152[,3]/subsampled_to) *100
names(p_152)[4] <- "Otu152_abund"

fusoviv <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient", "location", "Otu00472")]
fusoviv[,5] <- (fusoviv[,4]/4321)*100
names(fusoviv)[5] <- "Otu00472_relAbund"
```

@@

Next, we built a model to differentiate the proximal and distal lumenal samples. The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, AUC = `r round(cv10f_roc_muc$auc,3)`) compared to the proximal versus distal lumen (AUC = `r round(cv10f_roc_lum$auc, 3)`). The model that separated proximal and distal mucosal samples used `r length(2:ncol(aucrf_data_LRlumen))` features. OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B). The model used `r length(2:ncol(aucrf_data_LRbowel))` features to separate proximal and distal lumenal samples. Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C). This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also helped to differentiate between the two sides- such as _Anaerococcus_ and _Finegoldia_.

#### Bacterial OTUs associated with cancer are found in healthy individuals 
Given that specific bacterial species have been associated with colorectal cancer and IBD, we probed our sample set for these OTUs. Among our 100 samples, the most frequent sequence associated with the _Fusobacterium_ genus was OTU179, which aligns via BLASTn to _Fusobacterium nucleatum subsp animalis_ (100% over full length). This is the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. (@Castellarin2011). There were 14 samples with the _F. nucleatum subsp. animalis_ sequence. Four sequences were located in the proximal mucosa and three in the distal mucosal samples (Figure 6A). The second most frequent _Fusobacterium_ sequence was OTU472, which aligns with 99% identity to _F. varium_. In addition to _F. nucleatum_, _F. varium_ has been associated with IBD (@Lee2016). 4 study participants harbored _F. varium_ and the samples were split evenly between the proximal and distal mucosa (Figure 6B). OTU152 was similar to the members of the _Porphyromonas_ genus and the most frequent sequence in that OTU aligned to _Porphyromonas asacharolytica_ (99% over full length), another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Figure 6C). Among the samples that were positive for _P. asacharolytica_, the relative abundances for this OTU ranged from 0.01% - 16%. Thus, disease-associated OTUs could be found in our sample set of 20 healthy individuals.

@@

To get around the noisiness from a diverse set of samples, we built a Random Forest model to identify microbes specific to each side. For each comparison we identified top 5 OTUs that were strongly predictive of one site or another. Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis @Donaldson2015. For instance, the proximal mucosa contains the highest oxygen concentratons of the colon and  harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and aerobic _Psuedomonas_. The distal mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_. The model was less effective at classifying the proximal and distal lumenal contents, probably because the samples are arguably composed of the same bacteria but differ in water content.

We detected _F. nucleatum_ and _P. asacharolytica_ in 8 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans @Baxter2016 and have oncogenic properties in cell culture and in mice @Sears2014. Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the distal mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous distal mucosa @Albenberg2014. In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on proximal-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum @Mima2016.  Of the 8 (40%) individuals positive for _F. nucleatum_ in this study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa. Data examining bacterial biofilms on the mucosa of CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis @Dejea2014. Finally, _Fusobacterium_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model @Whitmore2014, @Flynn2016. Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of IBD or CRC subtypes in the proximal or distal colon.

The _Fusobacterium_ species _nucleatum_ and _varium_ have been commonly isolated from mucosal biopsies of patients with IBD @Lee2016. Laboratory experiments with these isolates have shown that disease-isolated _F. nucleatum_ are more invasive and stimulate more TNF-a production than strains from healthy individuals @Strauss2011, suggesting the bacteria may increase inflammation in the gut as well @Dharmani2011. _F. varium_ isolated from UC patients caused colonic ulcers in an experimental mouse model @Ohkusa2003. _F. varium_ was only detected in 3 of our study participants and two of those samples were isolated form the proximal mucosa (Figure 6B). Although there is low occurrance in our study, _F. varium_ is most commonly isolated from UC patient biopsies from the ileum or cecum @Ohkusa2002, suggesting this species may exhibit preference for the different environmental conditions of these gastrointestinal sites. Further work will assess how gut environment may select for species which then cause localized disease.

Specific comparisons of our findings to previously published gut biogeography studies are additionally confounded by the use of bowel preparation methods in most other studies. A rare report of a matched-colonoscopy study sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing @Shobar2016. This group found that mucosal and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site @Shobar2016. Bowel cleansing not only made the samples harder to distinguish, it resulted in decreases in diversity across sites. Further, the differences were not great enough to overcome interpersonal differences between subjects. Bowel preparation clearly induces bias into the microbes recovered from sampling the lumen or mucosa of a prepared bowel. Thus our findings in this study are strengthened by the lack of bowel preparation.

@@

#### Sample processing, sequencing and analysis
DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples. The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described @Kozich2013. Sequences were curated using the mothur software as described previously @Schloss2009. The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) @Wang2007 and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff. Sequencing and analysis of a mock community revealed the error rate to be X%. Samples were rarefied to 4231 sequences per sample in order to reduce uneven sampling bias.

Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur version <3.2?> @Schloss2009. ThetaYC distances were calculated to determine the dissimilarity between two samples. Random Forest classification models were built using the randomForest R package and resultant models were used to identify the OTUs that were most important for classifying each location @Liaw2002. To get species-level information about sequences of interest, sequences were aligned using blastn and the species name was only used if the identity score was >= 99%.

@@

####Figure 6
Location and abundance of cancer-associated OTUs. Relative abundance was calculated and plotted by sample site for each OTU of interest: (A) _Fusobacterium nucleatum subsp. animalis_ (B) _Fusobacterium varium_ and (C) _Porphyromonas asacharolytica_

### 2017-07-27 submission/manuscript.tex
fixed the bug
updated inline values and citations for v2

To identify OTUs that were distinct at each biogeographical site, we
constructed several Random Forest models trained using OTU abundances.
We built the first model to classify the lumen versus mucosal samples
for the proximal and distal sides, independently (Figure 4A). The
constructed model used ((Xopt)) features for the proximal and ((Xopt))
for the distal. The models performed well when classifying these samples
(0.8 and 1.0, respectively). The OTUs that were most predictive of each
site are identified by their greatest mean decrease in accuracy when
removed from the model. For distinguishing the proximal lumen and
mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
were differentially abundant (Figure 4B). The model classifying the
distal lumen and mucosa identified OTUs from \emph{Turicibacter},
\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
\emph{Anaerococcus} genera that could distinguish lumen from mucosal
samples (Figure 4C). These results indicate that there are fine
differences between the different sites of the colon, and that these can
be traced down to specific OTUs on each side.

Next, we built a model to differentiate the proximal and distal lumenal
samples. The model performed best when distinguishing the proximal
versus distal mucosa (Figure 5A, AUC = 0.912) compared to the proximal
versus distal lumen (AUC = 0.755). These models were able to explain
((X\%)) of the variance, respectively. OTUs that were differentially
abundant between the distal and proximal mucosa included members of the
\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B).
Differentially abundant OTUs of the proximal and distal lumen included
three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
and an \emph{Oscillibacter} OTU (Figure 5C). This analysis found that
some of the same OTUs that are distinct between the mucosa and lumen
also helped to differentiate between the two sides- such as
\emph{Anaerococcus} and \emph{Finegoldia}.

@@

\hypertarget{ref-Halfvarson2017}{}
3. \textbf{Halfvarson J}, \textbf{Brislawn CJ}, \textbf{Lamendella R},
\textbf{Vázquez-Baeza Y}, \textbf{Walters WA}, \textbf{Bramer LM},
\textbf{D\textquotesingleAmato M}, \textbf{Bonfiglio F},
\textbf{McDonald D}, \textbf{Gonzalez A}, \textbf{McClure EE},
\textbf{Dunklebarger MF}, \textbf{Knight R}, \textbf{Jansson JK}. 2017.
Dynamics of the human gut microbiome in inflammatory bowel disease.
Nature Microbiology \textbf{2}:17004.
doi:\href{https://doi.org/10.1038/nmicrobiol.2017.4}{10.1038/nmicrobiol.2017.4}.

>>>>

To identify OTUs that were distinct at each biogeographical site, we
constructed several Random Forest models trained using OTU abundances.
We built the first model to classify the lumen versus mucosal samples
for the proximal and distal sides, independently (Figure 4A). The
constructed model used 14 features for the proximal and 6 for the
distal. The models performed well when classifying these samples (0.8
and 1.0, respectively). The OTUs that were most predictive of each site
are identified by their greatest mean decrease in accuracy when removed
from the model. For distinguishing the proximal lumen and mucosa, OTUs
from the \emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and
two OTUs from the \emph{Enterobacteraceae} genera were differentially
abundant (Figure 4B). The model classifying the distal lumen and mucosa
identified OTUs from \emph{Turicibacter}, \emph{Finegoldia},
\emph{Peptoniphilus} and two OTUs from the \emph{Anaerococcus} genera
that could distinguish lumen from mucosal samples (Figure 4C). These
results indicate that there are fine differences between the different
sites of the colon, and that these can be traced down to specific OTUs
on each side.

Next, we built a model to differentiate the proximal and distal lumenal
samples. The model performed best when distinguishing the proximal
versus distal mucosa (Figure 5A, AUC = 0.9) compared to the proximal
versus distal lumen (AUC = 0.8). OTUs that were differentially abundant
between the distal and proximal mucosa included members of the
\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B).
Differentially abundant OTUs of the proximal and distal lumen included
three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
and an \emph{Oscillibacter} OTU (Figure 5C). This analysis found that
some of the same OTUs that are distinct between the mucosa and lumen
also helped to differentiate between the two sides- such as
\emph{Anaerococcus} and \emph{Finegoldia}.

@@

\hypertarget{ref-Halfvarson2017}{}
3. \textbf{Halfvarson J}, \textbf{Brislawn CJ}, \textbf{Lamendella R},
\textbf{Vazquez-Baeza Y}, \textbf{Walters WA}, \textbf{Bramer LM},
\textbf{DAmato M}, \textbf{Bonfiglio F}, \textbf{McDonald D},
\textbf{Gonzalez A}, \textbf{McClure EE}, \textbf{Dunklebarger MF},
\textbf{Knight R}, \textbf{Jansson JK}. 2017. Dynamics of the human gut
microbiome in inflammatory bowel disease. Nature Microbiology
\textbf{2}:17004.
doi:\href{https://doi.org/10.1038/nmicrobiol.2017.4}{10.1038/nmicrobiol.2017.4}.
